---
title: "Cross Validation"
subtitle: "Lab 10 - MATH 4322"
format: pdf
highlight-style: github
code-block-bg: true
code-block-border-left: "#31BAE9"
header-includes: 
- \newcommand{\bi}{\begin{itemize}}
- \newcommand{\ei}{\end{itemize}}
- \newcommand{\iset}{\setlength\itemsep{1em}}
- \usepackage{amsbsy}
- \usepackage{hyperref}
- \usepackage{multicol}
- \newcommand{\blue}[1]{\textcolor{blue}{#1}}
- \newcommand{\red}[1]{\textcolor{red}{#1}}
---

## Problem

We will use the data \blue{Weekly} in the `ISLR` package to predict \blue{Direction} using \blue{Lag1} and \blue{Lag2}.  

**Description**

Weekly percentage returns for the S&P 500 stock index between 1990 and 2010.  

\blue{Lag1} Percentage return for previous week  
\blue{Lag2} Percentage return for 2 weeks previous  
\blue{Direction} A factor with levels Down and Up indicating whether the market had a positive or negative return on a given week  

```{r}
#| echo: false
library(ISLR)
summary(Weekly)
```

\red{Question 1}:  Is this a regression or classification problem?  
  
    This is a classification problem because the response variable is binary.

\red{Question 2}:  Which is the correct model, linear regression or logistic regression?  


    Between the 2 models the logistic regression is the correct model.

\red{Question 3}:  Write out the equation of the correct model.  

$$\hat{Direction} = \frac{exp(\beta_0 + beta_1 x Lag1 + beta_2 x Lag2)}
{1 + exp(\beta_0 + beta_1 x Lag1 + beta_2 x Lag2)}$$


## Task 1

Fit a logistic regression model that predicts \blue{Direction} using \blue{Lag1} and \blue{Lag2} on half of the data. This is the **training** data.

\red{Question 4}:  In R what code do we use to separate the data into a train and test data.  

```{r}
#| echo: true
library(ISLR)
set.seed(100)
sample = sample(1:nrow(Weekly), nrow(Weekly)/2)
train = Weekly[sample,]
test = Weekly[-sample,]
```


\red{Question 5}: In R what code do would we use to get a model to predict direction based on lag1 and lag2?

Use this model to predict the direction of the first observation.
You can do this by predicting that the first observation
will go up if $P(Direction = \text{"Up"}|Lag1, Lag2) > 0.5$

\vspace{2em}

```{r}
#| echo: true
direction.glm = glm(Direction ~ Lag1 + Lag2, data = train, family = binomial)
direction.glm
```


\red{Question 6}:  What is the code to get a prediction of the first observation?  

```{r}
predict.glm(direction.glm, newdata = Weekly[1,], type = "response")
```

```{r}
Weekly[1,]$Direction
```


\red{Question 7}:  Would we predict the first observation to go up or down?  

    We would predict the first observation to go up.

\red{Question 8}:  Is the a correct prediction or miss classified?  

    The original observation is going Down, so this is misclassified.

We want to create a confusion matrix to determine the proportion of miss classified           observations.  This is called the *error rate*.

\red{Question 9}:  What is the code to create this confusion matrix?  

```{r}
#| echo: true
pred.train = predict.glm(direction.glm, type = "response")
pred.direction.train = ifelse(pred.train > 0.5, "Up", "Down")
(confmat.train = table(train$Direction, pred.direction.train))
```
\vspace{2em}

\red{Question 10}:  What is the error rate based on the training data?  

```{r}
(confmat.train[1, 2] + confmat.train[2, 1])/sum(confmat.train)
```


## Task 2  

Use this trained model to make a prediction from the data that we did not use.  This is the **test** data.

```{r}
pred.test = predict.glm(direction.glm, newdata = test)
pred.direction.test = ifelse(pred.test > 0.5, "Up", "Down")
(confmat.test = table(test$Direction, pred.direction.test))
```


\red{Question 11}:  What is the test error rate?  

```{r}
(confmat.test[1,2] + confmat.test[2,1])/sum(confmat.test)
```


## Task 3

We want to create a LOOCV test error rate for the whole data using a for loop.  

Write a for loop from $i = 1$ to $i = n$, where $n$ is the number of observations in the data set, that performs each of the following steps:

i. Fit a logistic regression model using all but the $i^{th}$ observation to predict Direction using Lag1 and Lag2.
ii. Compute the posterior probability of the market moving up
for the $i^{th}$ observation.
iii. Use the posterior probability for the $i^th$ observation in order to predict whether or not the market moves up.
iv. Determine whether or not an error was made in predicting
the direction for the $i^{th}$ observation. If an error was made,
then indicate this as a 1, and otherwise indicate it as a 0.  

```{r}
#| echo: true
#| warning: false
#| message: false
#| results: hide
loocv.err = NA
for (i in 1:nrow(Weekly)){
  sample = i
  fit.glm = glm(Direction ~ Lag1 + Lag2,data = Weekly[-sample,],
                family = binomial)
  pred.fit = predict.glm(fit.glm,Weekly[sample,],
                         type = "response")
  pred.direction = ifelse(pred.fit>0.5,"Up","Down")
  loocv.err[i] = (Weekly[sample,]$Direction != pred.direction)
}

```


\red{Question 12}:  Take the average of the $n$ numbers obtained in **iv** in order to obtain the LOOCV estimate for the test error. What is the value?  

\vspace{1em}

```{r}
#| echo: true
mean(loocv.err)
```



## Task 4 

We will use the `cv.glm` function to determine LOOCV estimate for the test error.  Since the response is binary, we will have to create a  **cost** function to determine what probability we want to use as a cut off for "Up".  



```{r}
#| echo: true
#| results: hide
library(boot)
#Since the response is a binary variable an appropriate cost function is
cost <- function(r, pi = 0) mean(abs(r-pi) > 0.5)
direction.glm =  glm(Direction ~ Lag1 + Lag2, data = Weekly,family = binomial)
cv.glm(Weekly,direction.glm,cost)$delta[1]
```

\red{Question 13}: Give the cross validation estimate from this method.  Compare this to the value in Task 3, is it the same, higher or lower?  

\vspace{1em}

    This is the same as previous value.


## Task 5

We will do a 10-fold cross validation

```{r}
#| echo: true
#| results: hide
set.seed(10)
cv.glm(Weekly,direction.glm,cost, K = 10)$delta[1]
```

\red{Question 14}:  Is this the same value as the loocv error?


    This is not the same value.

\red{Question 15}:  Repeat the `cv.glm` code again.  Do you get the same value?  

\vspace{2em}

```{r}
#| echo: true
#| results: hide
cv.glm(Weekly,direction.glm,cost, K = 10)$delta[1]
```

\vspace{1em}
    
    We do not get the same value as before.

\red{Question 16}:  What does the CV value represent in this scenario?  

    The CV is from the 10 samples, the average error rate.
